---
title: "R4DS_Ch02_MyNotes"
author: "LucyT"
format: html
editor: visual
---

# R for Data Science - Chapter 02 - My notes

Re-learning R and GitHub, while trying out Quarto.

## Set-up

**Packages**

Add packages:

```{r  include=FALSE}
list.of.packages <- c("tidyverse", "arrow", "babynames", "curl", "duckdb", "gapminder", 
    "ggrepel", "ggridges", "ggthemes", "hexbin", "janitor", "Lahman", 
    "leaflet", "maps", "nycflights13", "openxlsx", "palmerpenguins", 
    "repurrrsive", "tidymodels", "writexl")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages,repos = "http://cran.us.r-project.org")

```

Load in Libraries:

```{r  include=FALSE}
#define vector of packages
new_package <- c("tidyverse", "arrow", "babynames", "curl", "duckdb", "gapminder", 
    "ggrepel", "ggridges", "ggthemes", "hexbin", "janitor", "Lahman", 
    "leaflet", "maps", "nycflights13", "openxlsx", "palmerpenguins", 
    "repurrrsive", "tidymodels", "writexl")
lapply(new_package, library, character.only=TRUE)
```

##Chapter 2: Workflow: basics

Create new objects using assignment operator <-:
Combine multiple elements using a vector c():
Comments use #:
``` {r}
#Create a vecotr of primes
primes <- c(2, 3, 5, 7, 11, 13)
primes
```

###Calling functions

Call functions like this: function_name(arguement1 = value1, arguement2 = value2, ...)
Call the seq function:
``` {r}
seq(from = 1, to = 10)
seq(1, 10)
```

##Excercises:
1. Why does this code not work?
**Answer**: incorrect "i"

2. Tweak each of the following R commands so that they run correctly:
``` {r}
library(tidyverse)

ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

3. Press Alt + Shift + K. What happens?
**Answer**: shortcut keys menu
