---
title: "R4DS_MyNotes"
author: "LucyT"
format: html
editor: visual
---

# R for Data Science - My notes

Re-learning R and GitHub, while trying out Quarto.

## Set-up

**Quarto**

see <https://quarto.org>.

**Running Code**

Click **Render** to generate document. Embed code like this:

```{r}
1 + 1
```

Add `echo: false` option to disable printing of code.

**Packages**

Starting with the Tidyverse package, then add others as we progress:

```{r}
list.of.packages <- c("tidyverse", "arrow", "babynames", "curl", "duckdb", "gapminder", 
    "ggrepel", "ggridges", "ggthemes", "hexbin", "janitor", "Lahman", 
    "leaflet", "maps", "nycflights13", "openxlsx", "palmerpenguins", 
    "repurrrsive", "tidymodels", "writexl")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages,repos = "http://cran.us.r-project.org")

```

Load in Libraries to use the functions of the packages:

```{r}
#define vector of packages
new_package <- c("tidyverse", "arrow", "babynames", "curl", "duckdb", "gapminder", 
    "ggrepel", "ggridges", "ggthemes", "hexbin", "janitor", "Lahman", 
    "leaflet", "maps", "nycflights13", "openxlsx", "palmerpenguins", 
    "repurrrsive", "tidymodels", "writexl")
lapply(new_package, library, character.only=TRUE)
```

Updates to the Tidyverse package can be found using:

```{r}
tidyverse_update(repos = "http://cran.us.r-project.org")
```

##Chapter 1: Data Visualisation

Using primarily the **ggplot2** package, and the **palmerpenguins** data, with **ggthemes** to use a colorblind friendly colour palette.

To see the full data use **view**:

```{r}
view(penguins)
```

To see variables and first few observations use **glimpse**:

```{r}
glimpse(penguins)
```

### Use ggplot2 to add layers to a graph

Steps:

1.  Begin with the data
2.  Map variables to the Aesthetics of the plot
3.  Add geometric objects to represent the data (e.g. \_bar, \_line, \_boxplot, \_point)
4.  Modify the Aesthetic mapping
5.  Additional lines / regressions (e.g. linear model)
6.  Improve the labels
7.  Make colourblind friendly

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, 
                y = body_mass_g)
) +
  geom_point(
    mapping = aes(color = species,
                  shape = species)
  ) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", 
    y = "Body mass (g)",
    color = "Species", 
    shape = "Species"
  ) +
  scale_color_colorblind()
```

### Exercises

1.  How many rows are in penguins? How many columns?

```{r}
glimpse(penguins)
```
*Answer* 344 rows, 7 columns

2.  What does the bill_depth_mm variable in the penguins data frame describe?
```{r}
?penguins
```
*Answer* A number denoting bill length (millimeters)

3. Make a scatterplot of bill_depth_mm vs. bill_length_mm. Describe the relationship between these two variables.
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = bill_depth_mm, 
                y = bill_length_mm)
) +
  geom_point(
    mapping = aes(color = species,
                  shape = species)
) +
  labs(
    title = "Penguin bill length and depth",
    x = "Bill depth (mm)",
    y = "Bill length (mm)"
  )
```
*Answer* Positive correlation between bill length and depth.

4.What happens if you make a scatterplot of species vs. bill_depth_mm? What might be a better choice of geom?
```{r}
ggplot(
  data = penguins,
  mapping = aes(x = bill_depth_mm, 
                y = bill_length_mm)
) +
  geom_point(
    mapping = aes(color = species,
                  shape = species)
) +
  labs(
    title = "Penguin bill length and depth",
    x = "Bill depth (mm)",
    y = "Bill length (mm)"
  )
```


Why does the following give an error and how would you fix it?

ggplot(data = penguins) + geom_point()

What does the na.rm argument do in geom_point()? What is the default value of the argument? Create a scatterplot where you successfully use this argument set to TRUE.

Add the following caption to the plot you made in the previous exercise: “Data come from the palmerpenguins package.” Hint: Take a look at the documentation for labs().

Recreate the following visualization. What aesthetic should bill_depth_mm be mapped to? And should it be mapped at the global level or at the geom level?

A scatterplot of body mass vs. flipper length of penguins, colored by bill depth. A smooth curve of the relationship between body mass and flipper length is overlaid. The relationship is positive, fairly linear, and moderately strong.

Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

ggplot( data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island) ) + geom_point() + geom_smooth(se = FALSE)

Will these two graphs look different? Why/why not?

ggplot( data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g) ) + geom_point() + geom_smooth()

ggplot() + geom_point( data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g) ) + geom_smooth( data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g)
